---
title: "DREAM 4 Data Analysis"
author: "Vignette Author"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, message=FALSE}
library(CellNOptR)
library(CNORode)
library(dplyr)
library(tidyr)
library(bninfo)
library(artemis)
```

## Objectives

* Obtain contrast of strength and direction with prior edges
* Obtain a decline of L1 error
* Obtain a decline of causal entropy

The following works with the compressed DREAM 4 model.

```{r}
data_file <- system.file("extdata/datasets", "MD-LiverDREAM.csv",
                         package = "artemis")
model_file <- system.file("extdata/models", "PKN-LiverDREAM.sif", 
                          package = "artemis")
cnolist <- CNOlist(data_file)
model <- readSIF(model_file) # Need to expand possible and relationships.
mod_model <- indexFinder(cnolist, model) %>%
  {compressModel(model,.)}
plotModel(mod_model, cnolist)
```

Use an EM-ish approach to replace the missing values.
```{r}
library(neuralnet)
.data <- getSignals(cnolist)[["30"]]  %>%
  data.frame %>%
  {cbind(getStimuli(cnolist), .)}
p38_NAs <- is.na(.data$p38)
mek12_NAs <- is.na(.data$mek12)
.sim_data <- .data
fit <- neuralnet(mek12 ~ igf1+il1a+tgfa+tnfa+akt+erk12+ikb+jnk12+hsp27,
              data = na.omit(.sim_data[, -9]), hidden = 3, err.fct="sse", 
              linear.output =FALSE, likelihood = TRUE)
.sim_data$mek12[mek12_NAs] <- neuralnet::compute(fit, .sim_data[mek12_NAs, -c(9, 11)]) %$%
  net.result %>%
  as.numeric
for(i in 1:10){
  p38_trainer <- .sim_data[!p38_NAs,]
  p38_predictor <- .sim_data[p38_NAs, -9]
  p38_new_fit <- neuralnet(p38 ~ igf1+il1a+tgfa+tnfa+akt+erk12+ikb+jnk12+hsp27+mek12,
                data = p38_trainer, hidden = 5, err.fct="sse", 
                linear.output =FALSE, likelihood = TRUE) 
  p38_new <-  neuralnet::compute(p38_new_fit, p38_predictor) %$%
    net.result %>%
    as.numeric
  print(round(p38_new, 4))
  .sim_data$p38[p38_NAs] <- p38_new
  mek12_trainer <- .sim_data[!mek12_NAs, ]
  mek12_predictor <- .sim_data[mek12_NAs, -11]
  mek12_new_fit <- neuralnet(mek12 ~ igf1+il1a+tgfa+tnfa+akt+erk12+ikb+jnk12+hsp27+p38,
                data = mek12_trainer, hidden = 5, err.fct="sse", 
                linear.output =FALSE, likelihood = TRUE) 
  mek12_new <- neuralnet::compute(mek12_new_fit, mek12_predictor) %$%
    net.result %>%
    as.numeric
  #print(mek12_new)
  .sim_data$mek12[mek12_NAs] <- mek12_new
}  
cnolist2 <- cnolist
detach(package:neuralnet)
detach(package:MASS)
base_mat <- cnolist2@signals[["0"]]
base_mat[is.na(base_mat)] <- 0
cnolist2@signals[["0"]] <- base_mat
cnolist2@signals[["30"]] <- .sim_data %>% 
  select(akt, erk12, ikb, jnk12, p38, hsp27, mek12) %>%
  as.matrix
```

I coerce the values into a binary "on/off" variable using the ODE fit

```{r}
expanded_mod <- expandGates(mod_model, maxInputsPerGate=2)  #enable AND gates for better CPDs
ode_parameters <- expanded_mod %>%
  createLBodeContPars(LB_n = 1, LB_k = 0.1,
                      LB_tau = 0.01, UB_n = 5, UB_k = 0.9, 
                      UB_tau = 10, default_n = 3, default_k = 0.9, 
                      default_tau = 1, opt_n = TRUE, opt_k = TRUE,
                      opt_tau = TRUE, random = FALSE)
model_sim <- plotLBodeModelSim(cnolist2, expanded_mod, ode_parameters, 
                               timeSignals=seq(0,30,0.5))
inhib_mat <- getInhibitors(cnolist)
targets <- colnames(inhib_mat)
inhibs <- paste0("inh_", targets)
colnames(inhib_mat) <- inhibs
.data <- model_sim[[length(model_sim)]]  %>%
  apply(2, round) %>%
  data.frame %>%
  set_names(mod_model$namesSpecies) %>%  
  select(-tgfa, -igf1, -tnfa, -il1a) %>%
  {cbind(inhib_mat, .)} %>%
  {cbind(getStimuli(cnolist), .)} %>%
  filter(!is.na(map3k7)) %>%
  lapply(ordered) %>%
  data.frame 
gt <- empty.graph(setdiff(names(.data), inhibs))
arcs(gt) <-  model2sif(mod_model)[, c(1, 3)] %>%
  set_colnames(c("from", "to"))
gt_inh <- empty.graph(names(.data))
arcs(gt_inh) <- arcs(gt) %>%
  rbind(cbind(from = inhibs, to = targets))
net <- rbn(gt_inh, n = 10000, data = .data, fit = "bayes", iss = 1) %>%
  select(-inh_ikk, -inh_mek12, -inh_pi3k, -inh_p38) %>% 
  bn.fit(gt, data = .)
save(net, file = "inst/extdata/robjects/dream_net.rda")
```

Next I generate strength and edge orientation diagrams.  First I treat the distribution of all proteins as random.  Any causal information will be a product of v-structures.

Then I introduce the pathway knowledge that igf1, il1a, tgfa, and tnfa are receptors.  In the first case I suppose that there is low probability that these have incoming egdes, prior probability of these nodes having incoming edges is near 0.  In the second, I suppose I know the outcoming edges with near probability.  In call cases, I assume these are random.

```{r}
receptors <- c("igf1", "il1a", "tgfa", "tnfa")
bl <- lapply(receptors, 
             function(receptor){
               cbind(from = setdiff(nodes(net), receptor), to = receptor)
             }) %>%
             {do.call("rbind", .)} 
wl <- arcs(net)[1:10, ]
prior_edges <-  arcs2names(wl, directed = F)
starting_nets <- random.graph(nodes(net), 500, method = "ic-dag", burn.in = 100) 
boot <- lapply(starting_nets, function(start_net){
    tabu(rbn(net, 80), start = start_net, tabu = 50, 
                    score = "bde", iss = 1)
  }) %>%
  custom.strength(nodes(net), cpdag = FALSE) # Keep it false for now
plot_no_prior <- boot %>%
  reduce_averaging %>%
  mutate(
    edge = arcs2names(., directed_edges = F),
    entropy = orientation_entropy(.),
    detected = ifelse(edge %in% arcs2names(arcs(net), directed_edges = F), "T", "F"),
    prior = ifelse(edge %in% prior_edges, "prior", "not")
  ) %>%
  select(-direction) 
```

Now introducing the prior information:

```{r}
bl_arcs <- apply(bl, 1, paste0, collapse = "->")
bl_nets <- lapply(starting_nets, function(rand_net){
    new_net <- empty.graph(nodes(rand_net))
    new_arcs <- rand_net %>%
      arcs %>%
      data.frame %>%
      mutate(edges = arcs2names(arcs(rand_net), directed = TRUE)) %>%
      filter(!edges %in% bl_arcs) %>%
      select(from, to) %>%
      as.matrix
    arcs(new_net) <- new_arcs
    new_net
  })
boot_bl <- lapply(bl_nets, function(start_net){
    tabu(rbn(net, 80), start = start_net, tabu = 50, 
         blacklist = bl, score = "bde", iss = 1)
  }) %>%
  custom.strength(nodes(net), cpdag = FALSE)
plot_bl <- boot_bl %>%
  reduce_averaging %>%
  mutate(
    edge = arcs2names(., directed_edges = F),
    entropy = orientation_entropy(.),
    detected = plot_no_prior$detected,
    prior = plot_no_prior$prior) %>%
  select(-direction) 
```

```{r}
wl_nets <-lapply(bl_nets, function(rand_net){
  new_net <- empty.graph(nodes(rand_net))
  # use suppressWarnings to avoid warnings about duplicate arcs
  new_net <- suppressWarnings(`arcs<-`(new_net, value = rbind(arcs(rand_net), wl)))
  new_net
})
boot_wl <- lapply(wl_nets, function(start_net){
    tabu(rbn(net, 80), start = start_net, tabu = 50, 
         blacklist = bl, whitelist = wl,
         score = "bde", iss = 1)
  }) %>%
  custom.strength(nodes(net), cpdag = FALSE)
plot_wl <- boot_wl %>%
  reduce_averaging %>%
  mutate(
    edge = arcs2names(., directed_edges = F),
    entropy = orientation_entropy(.),
    detected = plot_no_prior$detected,
    prior = plot_no_prior$prior
  ) %>%
  select(-direction) 
edge_id <- order(order(rowMeans(cbind(plot_no_prior$strength,
                                      plot_bl$strength,
                                      plot_wl$strength)))) 
plot_no_prior$edge_id <- edge_id
plot_bl$edge_id <- edge_id
plot_wl$edge_id <- edge_id
```


Strength without prior knowledge
```{r}
plot_no_prior %>%
  ggplot(aes(x = edge_id, y = strength, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Strength of inferred edges, no prior knowledge")
```

Strength with low probability on incoming links to receptors.
```{r}
plot_bl %>%
  ggplot(aes(x = edge_id, y = strength, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Strength of inferred edges with low prob receptor edges")
```

Strength with high probability on outgoing links from receptors.
```{r}
plot_wl %>%
  ggplot(aes(x = edge_id, y = strength, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Strength of inferred edges with high prob receptor edges")
```

Now orientation entropy without prior knowledge:

```{r}
plot_no_prior %>%
  ggplot(aes(x = edge_id, y = entropy, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Orientation entropy without prior knowledge")
```

With low prob receptor edges:

```{r}
plot_bl %>%
  ggplot(aes(x = edge_id, y = entropy, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Orientation entropy with low prob receptor edges")
```

```{r}
plot_wl %>%
  ggplot(aes(x = edge_id, y = entropy, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Orientation entropy with high prob receptor edges")
```

# Repeating the analysis using prior probabilities instead of blacklist

```{r}
receptors <- c("igf1", "il1a", "tgfa", "tnfa")
bl <- lapply(receptors, 
             function(receptor){
               cbind(from = setdiff(nodes(net), receptor), to = receptor)
             }) %>%
             {do.call("rbind", .)} 
wl <- arcs(net)[1:10, ]
prior_edges <-  arcs2names(wl, directed = F)
starting_nets <- random.graph(nodes(net), 500, method = "ic-dag", burn.in = 100) 
boot <- lapply(starting_nets, function(start_net){
    tabu(rbn(net, 80), start = start_net, tabu = 50, 
                    score = "bde", iss = 40)
  }) %>%
  custom.strength(nodes(net), cpdag = FALSE) # Keep it false for now
plot_no_prior <- boot %>%
  reduce_averaging %>%
  mutate(
    edge = arcs2names(., directed_edges = F),
    entropy = orientation_entropy(.),
    detected = ifelse(edge %in% arcs2names(arcs(net), directed_edges = F), "T", "F"),
    prior = ifelse(edge %in% prior_edges, "prior", "not")
  ) %>%
  select(-direction) 
```

Now introducing the prior information:

```{r}
receptor_bl <- lapply(receptors,
                      function(receptor){
                        cbind(from = receptor, to = setdiff(receptors, receptor))
                      }) %>%
  {do.call("rbind", .)} %>%
  data.frame(stringsAsFactors = F) %>%
  mutate(prob = .Machine$double.eps * 100)
prior_no_incoming <- lapply(receptors, 
             function(receptor){
               cbind(from = setdiff(nodes(net), receptors), to = receptor)
             }) %>%
  {do.call("rbind", .)} %>%
  data.frame(stringsAsFactors = F) %>%
  mutate(prob = .46875 * .1)
prior_incoming <- lapply(receptors, 
             function(receptor){
               cbind(from = receptor, to = setdiff(nodes(net), receptors))
             }) %>%
  {do.call("rbind", .)} %>%
  data.frame(stringsAsFactors = F) %>%
  mutate(prob = .46875 * .9)

bl_prior <- rbind(receptor_bl, prior_no_incoming, prior_incoming) 
boot_bl <- lapply(starting_nets, function(start_net){
    tabu(rbn(net, 80), start = start_net, tabu = 50, 
         prior = "cs", beta = bl_prior, score = "bde", iss = 40)
  }) %>%
  custom.strength(nodes(net), cpdag = FALSE)
plot_bl <- boot_bl %>%
  reduce_averaging %>%
  mutate(
    edge = arcs2names(., directed_edges = F),
    entropy = orientation_entropy(.),
    detected = plot_no_prior$detected,
    prior = plot_no_prior$prior) %>%
  select(-direction) 
```

```{r}
prior_outgoing <- wl %>%
  data.frame %>%
  mutate(prob = .5 * .9)
full_prior <- rbind(receptor_bl, prior_no_incoming, prior_outgoing) 
boot_wl <- lapply(starting_nets, function(start_net){
    tabu(rbn(net, 80), start = start_net, tabu = 50, 
         prior = "cs", beta = full_prior,
         score = "bde", iss = 80)
  }) %>%
  custom.strength(nodes(net), cpdag = FALSE)
plot_wl <- boot_wl %>%
  reduce_averaging %>%
  mutate(
    edge = arcs2names(., directed_edges = F),
    entropy = orientation_entropy(.),
    detected = plot_no_prior$detected,
    prior = plot_no_prior$prior
  ) %>%
  select(-direction) 
edge_id <- order(order(rowMeans(cbind(plot_no_prior$strength,
                                      plot_bl$strength))))#,
#                                      plot_wl$strength)))) 
plot_no_prior$edge_id <- edge_id
plot_bl$edge_id <- edge_id
plot_wl$edge_id <- edge_id
```



Strength without prior knowledge
```{r}
plot_no_prior %>%
  ggplot(aes(x = edge_id, y = strength, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Strength of inferred edges, no prior knowledge")
```



Strength with low probability on incoming links to receptors.
```{r}
plot_bl %>%
  ggplot(aes(x = edge_id, y = strength, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Strength of inferred edges with low prob receptor edges")
```

Strength with high probability on outgoing links from receptors.
```{r}
plot_wl %>%
  ggplot(aes(x = edge_id, y = strength, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Strength of inferred edges with high prob receptor edges")
```

Now orientation entropy without prior knowledge:

```{r}
plot_no_prior %>%
  ggplot(aes(x = edge_id, y = entropy, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Orientation entropy without prior knowledge")
```

With low prob receptor edges:

```{r}
plot_bl %>%
  ggplot(aes(x = edge_id, y = entropy, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Orientation entropy with low prob receptor edges")
```

```{r}
plot_wl %>%
  ggplot(aes(x = edge_id, y = entropy, 
             colour = detected, shape = prior, size = prior)) +
  geom_point() +
  xlab("edge id") + 
  ggtitle("Orientation entropy with high prob receptor edges")
```





